---
title: "Virus Rating GWAS Analysis"
author: "Zachary Winn"
date: "2025-11-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is an analysis of the collected virus data associated CSU phenotypic data. In this analysis I do the following:

1.  Read in phenotypic data.
2.  Convert agronomic data to metric and transform virus data.
3.  Adjust for blocking factors within environment and calculate best linear unbiased estimates (BLUEs).
4.  Perform an adjustment across environments using the standard error of 1st stage BLUEs as weights on the residual.
5.  Read in genetic data (VCF).
6.  Perform a principle component analysis (PCA) on marker data.
7.  Fit a Q+K (population structure and relatedness) genome wide association study (GWAS) on the panel.
8.  Visualize the results.

## Read in phenotypic data and perform spatial adjustment (1st step analysis)

For normal and (transformed) ordinal traits, we use a spatial model in sommer to get BLUEs to be fed forward into a GWAS.

For ordinal virus ratings, we have employed an ranked, inverse normal transformation with a blom offset ($c=0.375$) to approximate a normal distribution. The transformation is as follows:

$$ 
INT(W_i) = \Phi^{-1} \left\{ \frac{rank(W_i) - c}{n + 1 - 2c}
$$

Where $W$ is the set of all original data points and $W_i$ is a single observation, $\Phi^{-1}$ is the inverse of the cumulative distribution function, $c$ is the blom offset ($c=0.375$), and $n$ is the total number of observations. 

For binomial traits (e.g., Julesburg 2025), raw data are retained to be fed forward into a binomial GWAS model where year_experiment_nursery (YEN) are treated as random (e.g., every environment is considered IID).

Here we do the following model to adjust within environment:

$$
y=\mu+G_i+rc_{jk}+\varepsilon_{ijk}
$$

Where y is the observation for the $i$-th genotype in the $j$-th row in the $k$-th column, $\mu$ is the population mean, $G_i$ is the fixed genotype effect. The term $rc_{jk}$ denotes the random row-by-column spatial interaction modeled as a two-dimensional penalized tensor-product of marginal B-splines. This spatial effect is defined as $rc = Z_s u_s$, where $Z_s$ is the design matrix constructed from the Kronecker product of row and column B-spline bases ($B_r \otimes B_c$), and $u_s$ is the vector of spline coefficients with $u_s \sim \mathcal{N}(0, \mathbf{I}\sigma^2_s)$. The residual error was assumed $\varepsilon_{ijk} \sim \mathcal{N}(0, \mathbf{I}\sigma^2_\varepsilon)$.

```{r single_env_adj}
# Library
library(ggplot2)
library(patchwork)

# Check dependencies
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("GENESIS", quietly = TRUE)) {
  BiocManager::install("GENESIS")
}

# Define the inverse normal transformation with blom offset function
INT_blom <- function(x) {
  # c = 3/8 (Blom offset)
  n <- sum(!is.na(x))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.375) / (n + 0.25))
}

# Read in pheno data
pheno <- read.csv("privatized_data.csv") |>
  dplyr::mutate(Y = ifelse(RESPONSE == "GY", Y * (67.25), # Conversion to Kg/H
    ifelse(RESPONSE == "TW", Y * (0.4536 / 0.3524), Y) # Conversion to Kg/Hl
  ))

# Check if the analysis needs to be ran
if (!file.exists("blues_single_environment.csv")) {
  # Get adjusted means object ready
  blues_se <- c()

  # Run adjustment within environment
  for (i in unique(pheno$YEN)) {
    # Make message
    print_message <- paste("### Analyzing YEN =", i, "###")
    border <- paste(rep("#", nchar(print_message)), collapse = "")
    print(border)
    print(print_message)
    print(border)

    # Pull data
    temp1 <- pheno[pheno$YEN == i, ]

    # For j in traits
    for (j in unique(temp1$RESPONSE)) {
      # Make message
      print_message <- paste("### Trait =", j, "###")
      border <- paste(rep("#", nchar(print_message)), collapse = "")
      print(border)
      print(print_message)
      print(border)

      # Pull trait
      temp2 <- temp1[temp1$RESPONSE == j, ]

      # Order row:col
      temp2 <- temp2[order(temp2$ROW, temp2$COLUMN), ]

      # Check if col 1 is col 1 if not set to 1
      temp2$ROW <- temp2$ROW |>
        as.factor() |>
        as.numeric()
      temp2$COLUMN <- temp2$COLUMN |>
        as.factor() |>
        as.numeric()
      temp2$Y <- as.numeric(temp2$Y)

      # Drop NA
      temp2 <- temp2 |> na.omit()

      # if temp2 has no data
      if (nrow(temp2) == 0) {
        remove(temp2)
        next
      }

      # Check type
      if (unique(temp2$DATA_TYPE) == "NORMAL") {
        # Run model in sommer
        temp3 <- try(sommer::mmes(
          fixed = Y ~ rID,
          random = ~ sommer::spl2Dc(ROW, COLUMN),
          rcov = ~units,
          data = temp2
        ))

        # If it fails toss the data
        if (class(temp3) == "try-error") {
          remove(temp2,temp3)
          next
        }

        # Calculate adjusted means
        temp4 <- sommer::predict.mmes(temp3, D = "rID")$pvals

        # Bind to adjusted means
        temp4$RESPONSE <- j
        temp4$YEN <- i
        temp4$DATA_TYPE <- "BLUE"
        blues_se <- rbind(blues_se, temp4)

        # Remove
        remove(temp2, temp3, temp4)
      } else if (unique(temp2$DATA_TYPE) == "ORDINAL") {
        # Transform y
        temp2$Y_t <- INT_blom(temp2$Y)

        # Run model in sommer
        temp3 <- try(sommer::mmes(
          fixed = Y_t ~ rID,
          random = ~ sommer::spl2Dc(ROW, COLUMN),
          rcov = ~units,
          data = temp2
        ))

        # If it fails toss the data
        if (class(temp3) == "try-error") {
          remove(temp2, temp3)
          next
        }

        # Calculate adjusted means
        temp4 <- sommer::predict.mmes(temp3, D = "rID")$pvals

        # Bind to adjusted means
        temp4$RESPONSE <- j
        temp4$YEN <- i
        temp4$DATA_TYPE <- "BLUE"
        blues_se <- rbind(blues_se, temp4)

        # Remove
        remove(temp2, temp3, temp4)
      } else if (unique(temp2$DATA_TYPE) == "BINOMIAL") {
        # Do nothing and just jam it into the blues dataframe
        temp3 <- data.frame(
          "rID" = temp2$rID,
          "predicted.value" = temp2$Y,
          "std.error" = NA,
          "RESPONSE" = j,
          "YEN" = i,
          "DATA_TYPE" = "BINOMIAL"
        )

        # Rbind
        blues_se <- rbind(blues_se, temp3)

        # Remove
        remove(temp2, temp3)
      }
    }
    # Remove
    remove(temp1)
  }

  # Write out
  write.csv(blues_se,
    "blues_single_environment.csv",
    row.names = FALSE
  )
} else {
  # Read in
  blues_se <- read.csv("blues_single_environment.csv")
}
```

## Calculate adjusted means across environment (2nd step analysis)

The BLUEs from the 1st stage analysis are now analyzed in a multiple-environment mixed model weighted by the precision of the first-stage estimates:

$$y_{ij} = \mu + G_i + E_j + \varepsilon_{ij}$$

where $y_{ij}$ is the adjusted mean (BLUE) for the $i$-th genotype in the $j$-th environment, $\mu$ is the overall mean, and $G_i$ is the effect of the $i$-th genotype. $E_j$ represents the random environment effect assuming $E_j \sim \mathcal{N}(0, \sigma^2_E)$. The residual error $\varepsilon_{ij}$ accounts for the heterogeneity of precision from the first stage, distributed as $\varepsilon_{ij} \sim \mathcal{N}(0, \sigma^2_{e}w_{ij}^{-1})$, where $w_{ij}$ represents the weights derived from the inverse of the squared standard errors ($1/SE_{ij}^2$) of the first-stage estimates.

```{r multi_env_adj}
# If exists
if (!file.exists("blues_multiple_environment.csv")) {
  # Create object for ME blues
  blues_me <- blues_se |>
    dplyr::distinct(rID)

  # Save multienvironment models
  models_me <- list()

  # Loop
  for (i in unique(blues_se$RESPONSE)) {
    # Make message
    print_message <- paste("### Analyzing Trait =", i, "###")
    border <- paste(rep("#", nchar(print_message)), collapse = "")
    print(border)
    print(print_message)
    print(border)

    # Pull dataframe
    temp1 <- blues_se[blues_se$RESPONSE == i & blues_se$DATA_TYPE == "BLUE", ]

    # Make message
    print_message <- paste("### System Time =", Sys.time(), "###")
    border <- paste(rep("#", nchar(print_message)), collapse = "")
    print(border)
    print(print_message)
    print(border)

    # Fit mmes gBLUP for GWAS
    temp2 <- sommer::mmes(
      fixed = predicted.value ~ rID,
      random = ~YEN,
      rcov = ~units,
      W = diag(temp1$std.error),
      data = temp1
    )

    # Make message
    print_message <- paste("### System Time =", Sys.time(), "###")
    border <- paste(rep("#", nchar(print_message)), collapse = "")
    print(border)
    print(print_message)
    print(border)

    # Predict
    temp3 <- sommer::predict.mmes(temp2, D = "rID")$pvals

    # Rename
    colnames(temp3)[2:ncol(temp3)] <- paste(i, colnames(temp3)[2:ncol(temp3)], sep = "_")

    # Left join
    blues_me <- blues_me |> dplyr::left_join(temp3, by = "rID")

    # Save model
    models_me[[i]]$data <- temp1
    models_me[[i]]$model <- temp2

    # Remove
    remove(temp1, temp2, temp3)
  }

  # Write out adj means
  write.csv(blues_me,
    "blues_multiple_environment.csv",
    row.names = FALSE
  )

  # Make object for correlation visualization
  blues_me_corr <- blues_me |>
    dplyr::select(rID, GY_predicted.value, TW_predicted.value, Virus_predicted.value) |>
    dplyr::rename(
      `Grain Yield (Kg/Ha)` = GY_predicted.value,
      `Test Weight (Kg/Hl)` = TW_predicted.value,
      `Virus Ratings (Transformed)` = Virus_predicted.value
    ) |>
    tidyr::drop_na()

  # Now look at correlation among BLUEs
  jpeg(
    filename = "blues_multiple_environment_pairs_plotjpg",
    width = 8,
    height = 8,
    units = "in",
    res = 320
  )

  psych::pairs.panels(blues_me_corr[, 2:4],
    hist.col = "gray",
    stars = TRUE,
    lm = TRUE,
    line.col = "red",
    density = TRUE
  )

  dev.off()
} else {
  # Read data
  blues_me <- read.csv("blues_multiple_environment.csv")
}
```

## Read in genetic data and perform for GWAS (3rd step analysis)

Marker effects were estimated using a generalized least squares model where the response variable was the Best Linear Unbiased Estimator for each genotype derived from the second-stage analysis. Unlike the preceding stage, this analysis assumed homoscedastic errors and did not incorporate observation weights. The model for the $i$-th genotype and $j$-th marker was defined as:

$$y_i = \mu + S_{ij}\alpha_j + u_i + \varepsilon_i$$

where $y_i$ is the BLUE for the $i$-th genotype, $\mu$ is the population intercept, and $S_{ij}$ represents the marker dosage (coded 0, 1, or 2) for the $j$-th SNP with fixed effect $\alpha_j$. The random term $u_i$ accounts for cryptic relatedness and population structure, distributed as $u \sim \mathcal{N}(0, \mathbf{K}\sigma^2_u)$, where $\mathbf{K}$ is the realized genomic relationship matrix. The residual error was assumed independent and identically distributed as $\varepsilon_i \sim \mathcal{N}(0, \sigma^2_e)$.

```{r gwas_continuous}
# Read in data
geno <- gaston::read.vcf("virus_gwas_2025_production_final.vcf.gz", convert.chr = FALSE)

# Make a marker map
marker_map <- geno@snps |>
  dplyr::rename(
    chr_wheat = chr,
    snp_id = id
  ) |>
  dplyr::mutate(
    chr = as.numeric(as.factor(chr_wheat)),
    variant.id = c(1:nrow(geno@snps))
  ) |>
  dplyr::select(
    chr,
    chr_wheat,
    variant.id,
    snp_id,
    pos
  ) |>
  dplyr::mutate(
    chr = as.integer(chr),
    variant.id = as.integer(variant.id),
    pos = as.integer(pos)
  )

# Subset phenotypic infomration
pheno_file <- blues_me |>
  dplyr::filter(rID %in% geno@ped$id) |>
  dplyr::select(rID, GY_predicted.value, TW_predicted.value, Virus_predicted.value) |>
  dplyr::rename(
    GY = GY_predicted.value,
    TW = TW_predicted.value,
    Virus = Virus_predicted.value
  )

# Subset genotypic information
geno_file <- gaston::as.matrix(geno)
geno_file <- geno_file[rownames(geno_file) %in% pheno_file$rID, ]

# Now make a line map
line_map <- data.frame(
  rID = rownames(geno_file),
  scanID = 1:length(rownames(geno_file))
)

# Now bind the phenotype file with the scanID and genotype rownames
rownames(geno_file) <- line_map$scanID
pheno_file <- line_map |>
  dplyr::left_join(pheno_file, by = "rID")

# Make geontype data object for GENISIS
geno_genisis <- GWASTools::MatrixGenotypeReader(
  genotype = t(geno_file),
  snpID = marker_map$variant.id,
  chromosome = marker_map$chr,
  position = marker_map$pos,
  scanID = line_map$scanID
)

# Now make the genotype data
geno_data_genisis <- GWASTools::GenotypeData(geno_genisis)

# Calculate kinship matrix
kinship_file <- sommer::A.mat(geno_file - 1)

# Make message for system time
print_message <- paste("### System Time =", Sys.time(), "###")
border <- paste(rep("#", nchar(print_message)), collapse = "")
print(border)
print(print_message)
print(border)

# Do PCA
pcs_geno <- prcomp(geno_file,
  center = TRUE,
  scale. = TRUE
)

# Make message for system time
print_message <- paste("### System Time =", Sys.time(), "###")
border <- paste(rep("#", nchar(print_message)), collapse = "")
print(border)
print(print_message)
print(border)

# Pull percent variance
pve_geno <- summary(pcs_geno)$importance

# Create labels with variance explained
pve_labs <- paste0(
  "PC",
  1:ncol(pve_geno),
  " (",
  round(pve_geno[2, 1:ncol(pve_geno)] * 100, 2),
  "%)"
)

# Now pull PCs
pcs_geno <- pcs_geno$x

# Make dataframe
pcs_geno <- as.data.frame(pcs_geno)

# Generate and stitch plots
pcs_geno_image <- ggplot(pcs_geno, aes(PC1, PC2)) +
  geom_point() +
  theme_bw() +
  labs(x = pve_labs[1], y = pve_labs[2]) +
  ggplot(pcs_geno, aes(PC1, PC3)) +
  geom_point() +
  theme_bw() +
  labs(x = pve_labs[1], y = pve_labs[3]) +
  ggplot(pcs_geno, aes(PC2, PC3)) +
  geom_point() +
  theme_bw() +
  labs(x = pve_labs[2], y = pve_labs[3]) +
  patchwork::plot_annotation(title = "Principal Component Analysis of Genetic Data")

# Save image
ggsave(
  filename = "geno_pca_comparison_continuous.jpg",
  pcs_geno_image,
  width = 15,
  height = 5,
  units = "in",
  dpi = 320
)

# Make a new covar file for the final scanannot file
covar_file_scanAnnot <- pcs_geno |>
  tibble::rownames_to_column(var = "scanID")

# Pull first 10 PCs
covar_file_scanAnnot <- as.data.frame(covar_file_scanAnnot[, 1:11])
covar_file_scanAnnot$scanID <- as.integer(covar_file_scanAnnot$scanID)

# Make scanannot
scanAnnot <- pheno_file |>
  dplyr::left_join(covar_file_scanAnnot, by = "scanID") |>
  dplyr::select(scanID, PC1, PC2, PC3, GY, TW, Virus) |>
  GWASTools::ScanAnnotationDataFrame()

# Remove
remove(covar_file_scanAnnot)

# Plot function
plot_gwas <- function(df,
                      chr = "chr_wheat",
                      pos = "pos",
                      y = "neg_log_pval",
                      threshold = NULL,
                      title = NULL) {
  # We assume the input is sorted by chromosome and position
  plot_data <- data.frame(CHR = df[[chr]], POS = df[[pos]], Y = df[[y]])
  plot_data$CHR <- factor(plot_data$CHR, levels = unique(plot_data$CHR))

  # tapply gets the max POS per chromosome; cumsum creates the running total
  chr_max <- tapply(plot_data$POS, plot_data$CHR, max)
  offsets <- c(0, cumsum(as.numeric(chr_max))[-length(chr_max)])

  # Map offsets to the dataframe using the factor integer indices
  plot_data$BP_CUM <- plot_data$POS + offsets[as.numeric(plot_data$CHR)]

  # Calculate X-axis breaks (centers of each chromosome)
  axis_breaks <- tapply(plot_data$BP_CUM, plot_data$CHR, mean)

  # Generate plot
  plot_obj <- ggplot2::ggplot(plot_data, ggplot2::aes(x = BP_CUM, y = Y)) +
    ggplot2::geom_point(ggplot2::aes(color = CHR), alpha = 0.8, size = 1.5) +
    ggplot2::scale_color_manual(values = rep(c("grey", "black"), nlevels(plot_data$CHR))) +
    ggplot2::scale_x_continuous(label = names(axis_breaks), breaks = axis_breaks) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      legend.position = "none",
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor.x = ggplot2::element_blank()
    ) +
    ggplot2::labs(x = "Chromosome", y = expression(-log[10](italic(P))))

  # if threshold != NULL
  if (!is.null(threshold)) {
    plot_obj <- plot_obj + ggplot2::geom_hline(
      yintercept = threshold,
      color = "red",
      linewidth = 1,
      linetype = "dashed"
    )
  }

  # if title != NULL
  if (!is.null(title)) {
    plot_obj <- plot_obj + ggplot2::labs(title = title)
  }

  # Return
  return(plot_obj)
}

# Make results object
GWAS_results_continuous <- c()

# Loop through traits
for (i in c("GY", "TW", "Virus")) {
  # Make message for system time
  print_message <- paste("### Analyzing Trait =", i, "###")
  border <- paste(rep("#", nchar(print_message)), collapse = "")
  print(border)
  print(print_message)
  print(border)
    
  # Make message for system time
  print_message <- paste("### System Time =", Sys.time(), "###")
  border <- paste(rep("#", nchar(print_message)), collapse = "")
  print(border)
  print(print_message)
  print(border)
  
  # Run Null Model for normal response
  temp1 <- GENESIS::fitNullModel(scanAnnot,
    outcome = i,
    covars = c("PC1", "PC2", "PC3"),
    cov.mat = kinship_file
  )

  # Genotype Block Iterator for GWAS
  temp2 <- GWASTools::GenotypeBlockIterator(geno_data_genisis)

  # Now run GWAS with null model
  temp3 <- suppressWarnings(GENESIS::assocTestSingle(temp2, null.model = temp1))

  # Now bind in true chromosome names and such
  temp4 <- marker_map |>
    dplyr::mutate(chr = as.character(chr)) |>
    dplyr::left_join(temp3, by = c("chr", "variant.id", "pos")) |>
    dplyr::mutate(
      neg_log_pval = -log10(Score.pval),
      trait = i
    ) |>
    dplyr::select(
      trait,
      chr,
      chr_wheat,
      variant.id,
      snp_id,
      pos,
      n.obs,
      freq,
      MAC,
      Score,
      Score.SE,
      Score.Stat,
      Score.pval,
      neg_log_pval,
      Est,
      Est.SE,
      PVE
    ) |>
    dplyr::mutate(break_bft = ifelse(neg_log_pval >= -log10(0.05/nrow(temp3)),
      1,
      0
    ))

  # Now plot
  temp5 <- plot_gwas(temp4,
    threshold = -log10(0.05/nrow(temp3))
  )

  # Now save image
  ggsave(
    filename = paste("gwas_manhattan_plot_", i, ".jpg", sep = ""),
    plot = temp5,
    width = 11,
    height = 8.5,
    units = "in",
    dpi = 320
  )

  # Now bind results
  GWAS_results_continuous <- rbind(GWAS_results_continuous, temp4)

  # Remove
  remove(temp1, temp2, temp3, temp4, temp5)
  
  # Make message for system time
  print_message <- paste("### System Time =", Sys.time(), "###")
  border <- paste(rep("#", nchar(print_message)), collapse = "")
  print(border)
  print(print_message)
  print(border)  
}

# Write out results
write.csv(GWAS_results_continuous,
          "gwas_results_continuous.csv",
          row.names = FALSE)
```

## Binomial GWAS

Unadjusted observations of disease presence/absence were analyzed similarly in a GWAS.

```{r gwas_binomial}
# Pull observations for binomial ratings
binomial_ratings <- blues_se |>
  dplyr::filter(DATA_TYPE == "BINOMIAL") |>
  dplyr::select(YEN, rID, predicted.value) |>
  dplyr::rename(virus_binomial = predicted.value)

# Get the mode of each rating
binomial_ratings_consensus <- c()

# Loop through ratings
for (i in unique(binomial_ratings$rID)) {
  # Pull data
  temp1 <- binomial_ratings |>
    dplyr::filter(rID == i) |>
    dplyr::select(rID, virus_binomial)

  # Check if there is only one rating
  if (nrow(temp1) == 1) {
    # Bind
    binomial_ratings_consensus <- rbind(binomial_ratings_consensus, temp1)

    # Remove data
    remove(temp1)

    # Go to next in i
    next
  }

  # Now take table
  temp2 <- table(temp1$virus_binomial)

  # If table only has one rating
  if (length(temp2) == 1) {
    # Get single rating
    temp1 <- temp1 |>
      dplyr::distinct()

    # Bind
    binomial_ratings_consensus <- rbind(binomial_ratings_consensus, temp1)

    # Remove data
    remove(temp1, temp2)

    # Go to next in i
    next
  }

  # Take maximum
  temp3 <- names(temp2[temp2 == max(temp2)]) |> as.numeric()

  # If there is a tie
  if (length(temp3) > 1 | length(temp3) < 1) {
    # remove
    remove(temp1, temp2, temp3)

    # Skip
    next
  } else {
    # Make dataframe
    temp4 <- data.frame(
      rID = i,
      virus_binomial = temp3
    )

    # Bind
    binomial_ratings_consensus <- rbind(binomial_ratings_consensus, temp4)

    # Remove
    remove(temp1, temp2, temp3, temp4)
  }
}

# Make a marker map
marker_map <- geno@snps |>
  dplyr::rename(
    chr_wheat = chr,
    snp_id = id
  ) |>
  dplyr::mutate(
    chr = as.numeric(as.factor(chr_wheat)),
    variant.id = c(1:nrow(geno@snps))
  ) |>
  dplyr::select(
    chr,
    chr_wheat,
    variant.id,
    snp_id,
    pos
  ) |>
  dplyr::mutate(
    chr = as.integer(chr),
    variant.id = as.integer(variant.id),
    pos = as.integer(pos)
  )

# Subset phenotypic infomration
pheno_file <- binomial_ratings_consensus |>
  dplyr::filter(rID %in% geno@ped$id) |>
  dplyr::select(rID, virus_binomial) |>
  dplyr::rename(Virus = virus_binomial)

# Subset genotypic information
geno_file <- gaston::as.matrix(geno)
geno_file <- geno_file[rownames(geno_file) %in% pheno_file$rID, ]

# Now make a line map
line_map <- data.frame(
  rID = rownames(geno_file),
  scanID = 1:length(rownames(geno_file))
)

# Now bind the phenotype file with the scanID and genotype rownames
rownames(geno_file) <- line_map$scanID
pheno_file <- line_map |>
  dplyr::left_join(pheno_file, by = "rID")

# Make geontype data object for GENISIS
geno_genisis <- GWASTools::MatrixGenotypeReader(
  genotype = t(geno_file),
  snpID = marker_map$variant.id,
  chromosome = marker_map$chr,
  position = marker_map$pos,
  scanID = line_map$scanID
)

# Now make the genotype data
geno_data_genisis <- GWASTools::GenotypeData(geno_genisis)

# Calculate kinship matrix
kinship_file <- sommer::A.mat(geno_file - 1)

# Make message for system time
print_message <- paste("### System Time =", Sys.time(), "###")
border <- paste(rep("#", nchar(print_message)), collapse = "")
print(border)
print(print_message)
print(border)

# Do PCA
pcs_geno <- prcomp(geno_file,
  center = TRUE,
  scale. = TRUE
)

# Make message for system time
print_message <- paste("### System Time =", Sys.time(), "###")
border <- paste(rep("#", nchar(print_message)), collapse = "")
print(border)
print(print_message)
print(border)

# Pull percent variance
pve_geno <- summary(pcs_geno)$importance

# Create labels with variance explained
pve_labs <- paste0(
  "PC",
  1:ncol(pve_geno),
  " (",
  round(pve_geno[2, 1:ncol(pve_geno)] * 100, 2),
  "%)"
)

# Now pull PCs
pcs_geno <- pcs_geno$x

# Make dataframe
pcs_geno <- as.data.frame(pcs_geno)

# Generate and stitch plots
pcs_geno_image <- ggplot(pcs_geno, aes(PC1, PC2)) +
  geom_point() +
  theme_bw() +
  labs(x = pve_labs[1], y = pve_labs[2]) +
  ggplot(pcs_geno, aes(PC1, PC3)) +
  geom_point() +
  theme_bw() +
  labs(x = pve_labs[1], y = pve_labs[3]) +
  ggplot(pcs_geno, aes(PC2, PC3)) +
  geom_point() +
  theme_bw() +
  labs(x = pve_labs[2], y = pve_labs[3]) +
  patchwork::plot_annotation(title = "Principal Component Analysis of Genetic Data")

# Save image
ggsave(
  filename = "geno_pca_comparison_binomial.jpg",
  pcs_geno_image,
  width = 15,
  height = 5,
  units = "in",
  dpi = 320
)

# Make a new covar file for the final scanannot file
covar_file_scanAnnot <- pcs_geno |>
  tibble::rownames_to_column(var = "scanID")

# Pull first 10 PCs
covar_file_scanAnnot <- as.data.frame(covar_file_scanAnnot[, 1:11])
covar_file_scanAnnot$scanID <- as.integer(covar_file_scanAnnot$scanID)

# Make scanannot
scanAnnot <- pheno_file |>
  dplyr::left_join(covar_file_scanAnnot, by = "scanID") |>
  dplyr::select(scanID, PC1, PC2, PC3, Virus) |>
  GWASTools::ScanAnnotationDataFrame()

# Remove
remove(covar_file_scanAnnot)

# Make results object
GWAS_results_binomial <- c()

# Loop through traits
for (i in c("Virus")) {
  # Make message for system time
  print_message <- paste("### Analyzing Trait =", i, "###")
  border <- paste(rep("#", nchar(print_message)), collapse = "")
  print(border)
  print(print_message)
  print(border)
    
  # Make message for system time
  print_message <- paste("### System Time =", Sys.time(), "###")
  border <- paste(rep("#", nchar(print_message)), collapse = "")
  print(border)
  print(print_message)
  print(border)
  
  # Run Null Model for binomial response
  temp1 <- GENESIS::fitNullModel(scanAnnot,
    outcome = i,
    covars = c("PC1", "PC2", "PC3"),
    cov.mat = kinship_file,
    family = "binomial"
  )

  # Genotype Block Iterator for GWAS
  temp2 <- GWASTools::GenotypeBlockIterator(geno_data_genisis)

  # Now run GWAS with null model
  temp3 <- suppressWarnings(GENESIS::assocTestSingle(temp2, null.model = temp1))

  # Now bind in true chromosome names and such
  temp4 <- marker_map |>
    dplyr::mutate(chr = as.character(chr)) |>
    dplyr::left_join(temp3, by = c("chr", "variant.id", "pos")) |>
    dplyr::mutate(
      neg_log_pval = -log10(Score.pval),
      trait = i
    ) |>
    dplyr::select(
      trait,
      chr,
      chr_wheat,
      variant.id,
      snp_id,
      pos,
      n.obs,
      freq,
      MAC,
      Score,
      Score.SE,
      Score.Stat,
      Score.pval,
      neg_log_pval,
      Est,
      Est.SE,
      PVE
    ) |>
    dplyr::mutate(break_bft = ifelse(neg_log_pval >= -log10(0.05/nrow(temp3)),
      1,
      0
    ))

  # Now plot
  temp5 <- plot_gwas(temp4,
    threshold = -log10(0.05/nrow(temp3))
  )

  # Now save image
  ggsave(
    filename = paste("gwas_manhattan_plot_", i, "_binomial.jpg", sep = ""),
    plot = temp5,
    width = 11,
    height = 8.5,
    units = "in",
    dpi = 320
  )

  # Now bind results
  GWAS_results_binomial <- rbind(GWAS_results_binomial, temp4)

  # Remove
  remove(temp1, temp2, temp3, temp4, temp5)
  
  # Make message for system time
  print_message <- paste("### System Time =", Sys.time(), "###")
  border <- paste(rep("#", nchar(print_message)), collapse = "")
  print(border)
  print(print_message)
  print(border)  
}

# Write out results
write.csv(GWAS_results_binomial,
          "gwas_results_binomial.csv",
          row.names = FALSE)
```

## Legacy Code

Here is some code to test the analysis using rrBLUP/statgenGWAS to check consistency between models.

```{r legacy_code, eval=FALSE, include=FALSE}
# # Read in data
# geno <- gaston::read.vcf("virus_gwas_2025_production_final.vcf.gz", convert.chr = FALSE)
#
# #### statgenGWAS ####
# # Make the marker map
# marker_map <- geno@snps
# rownames(marker_map) <- marker_map[, 2]
# marker_map <- marker_map[, c(1, 4)]
#
# # Make the marker matrix
# marker_mat <- gaston::as.matrix(geno)
#
# # Read in phenotypic file
# pheno_file <- blues_me[, -grep("std.err", colnames(blues_me))]
#
# # Subset phenotyped individuals to only have individuals featured in marker matrix
# pheno_file <- pheno_file[pheno_file$rID %in% rownames(marker_mat), ]
#
# # Subset marker matrix to only have phenotyped invdividuals
# marker_mat <- marker_mat[rownames(marker_mat) %in% blues_me$rID, ]
#
# # Calculate kinship matrix
# kinship_mat <- sommer::A.mat(marker_mat - 1)
#
# # Make message for system time
# print_message <- paste("### System Time =", Sys.time(), "###")
# border <- paste(rep("#", nchar(print_message)), collapse = "")
# print(border)
# print(print_message)
# print(border)
#
# # Do PCA
# pcs_geno <- prcomp(marker_mat,
#   center = TRUE,
#   scale. = TRUE
# )
#
# # Make message for system time
# print_message <- paste("### System Time =", Sys.time(), "###")
# border <- paste(rep("#", nchar(print_message)), collapse = "")
# print(border)
# print(print_message)
# print(border)
#
# # Pull percent variance
# pve_geno <- summary(pcs_geno)$importance
#
# # Create labels with variance explained
# pve_labs <- paste0(
#   "PC",
#   1:ncol(pve_geno),
#   " (",
#   round(pve_geno[2, 1:ncol(pve_geno)] * 100, 2),
#   "%)"
# )
#
# # Now pull PCs
# pcs_geno <- pcs_geno$x
#
# # Make dataframe
# pcs_geno <- as.data.frame(pcs_geno)
#
# # Generate and stitch plots
# pcs_geno_image <- ggplot(pcs_geno, aes(PC1, PC2)) +
#   geom_point() +
#   theme_bw() +
#   labs(x = pve_labs[1], y = pve_labs[2]) +
#   ggplot(pcs_geno, aes(PC1, PC3)) +
#   geom_point() +
#   theme_bw() +
#   labs(x = pve_labs[1], y = pve_labs[3]) +
#   ggplot(pcs_geno, aes(PC2, PC3)) +
#   geom_point() +
#   theme_bw() +
#   labs(x = pve_labs[2], y = pve_labs[3]) +
#   patchwork::plot_annotation(title = "Principal Component Analysis of Genetic Data")
#
# # Save image
# ggsave(
#   filename = "geno_pca_comparison.jpg",
#   pcs_geno_image,
#   width = 15,
#   height = 5,
#   units = "in",
#   dpi = 320
# )
#
# # Make a covariate file with just lines in phenotype (just to make sure)
# covar_file <- pcs_geno[rownames(pcs_geno) %in% pheno_file$rID, ]
#
# # Make sure first column is titled "genotype"
# colnames(pheno_file)[1] <- "genotype"
#
# # Make gdata object for statgenGWAS
# gdata_obj <- statgenGWAS::createGData(
#   geno = marker_mat,
#   map = marker_map,
#   kin = kinship_mat,
#   pheno = pheno_file,
#   covar = covar_file
# )
#
# # Make message for system time
# print_message <- paste("### System Time =", Sys.time(), "###")
# border <- paste(rep("#", nchar(print_message)), collapse = "")
# print(border)
# print(print_message)
# print(border)
#
# # Run GWAS on all traits
# GWAS_Results <- statgenGWAS::runSingleTraitGwas(
#   gData = gdata_obj,
#   traits = colnames(pheno_file)[2:ncol(pheno_file)],
#   covar = c("PC1", "PC2", "PC3"),
#   kin = gdata_obj$kinship
# )
#
# # Make message for system time
# print_message <- paste("### System Time =", Sys.time(), "###")
# border <- paste(rep("#", nchar(print_message)), collapse = "")
# print(border)
# print(print_message)
# print(border)
#
# # Quick plot
# gy_gwas_img <- plot(GWAS_Results,
#   plotType = "manhattan",
#   trait = "GY_predicted.value",
#   title = "Grain Yield (Kg/Ha)"
# )
# tw_gwas_img <- plot(GWAS_Results,
#   plotType = "manhattan",
#   trait = "TW_predicted.value",
#   title = "Test Weight (Kg/Hl)"
# )
# virus_gwas_img <- plot(GWAS_Results,
#   plotType = "manhattan",
#   trait = "Virus_predicted.value",
#   title = "Virus (Transformed)"
# )
#
# # Write images
# ggsave("gwas_image_gy.jpg",
#   plot = gy_gwas_img,
#   width = 20,
#   height = 10,
#   units = "in",
#   dpi = 320
# )
# ggsave("gwas_image_tw.jpg",
#   plot = tw_gwas_img,
#   width = 20,
#   height = 10,
#   units = "in",
#   dpi = 320
# )
# ggsave("gwas_image_virus.jpg",
#   plot = virus_gwas_img,
#   width = 20,
#   height = 10,
#   units = "in",
#   dpi = 320
# )
#
# # Write out results
# write.csv(GWAS_Results$GWAResult$pheno_file,
#   "gwas_results_full.csv",
#   row.names = FALSE
# )

# #### rrBLUP ####
# # Make marker matrix
# marker_map <- geno@snps
# marker_map <- marker_map[, c(2,1,4)]
# marker_mat <- t(gaston::as.matrix(geno)-1) |>
#   as.data.frame() |>
#   tibble::rownames_to_column(var = "id") |>
#   as.data.frame()
# marker_mat <- marker_map |>
#   dplyr::left_join(marker_mat, by = "id")
#
# # Make phenotype file
# pheno_file <- blues_me[, -grep("std.err", colnames(blues_me))]
# pheno_file <- pheno_file[pheno_file$rID %in% colnames(marker_mat), ]
#
# # Make sure that the individuals who are in marker matrix are in phenotypic matrix
# lines <- colnames(marker_mat)[colnames(marker_mat) %in% pheno_file$rID]
# marker_mat <- marker_mat[,c(colnames(marker_mat)[1:3],
#                             colnames(marker_mat)[colnames(marker_mat) %in% lines])]
#
# # Plot function
# plot_gwas <- function(df, chr = "chr", pos = "position", y = "-log(p)") {
#
#   # Create a local subset with standardized names to avoid tidy-eval dependencies
#   # We assume the input is sorted by chromosome and position
#   plot_data <- data.frame(CHR = df[[chr]], POS = df[[pos]], Y = df[[y]])
#   plot_data$CHR <- factor(plot_data$CHR, levels = unique(plot_data$CHR))
#
#   # Calculate cumulative position offsets using base R
#   # tapply gets the max POS per chromosome; cumsum creates the running total
#   chr_max <- tapply(plot_data$POS, plot_data$CHR, max)
#   offsets <- c(0, cumsum(as.numeric(chr_max))[-length(chr_max)])
#
#   # Map offsets to the dataframe using the factor integer indices
#   plot_data$BP_CUM <- plot_data$POS + offsets[as.numeric(plot_data$CHR)]
#
#   # Calculate X-axis breaks (centers of each chromosome)
#   axis_breaks <- tapply(plot_data$BP_CUM, plot_data$CHR, mean)
#
#   # Generate plot
#   ggplot2::ggplot(plot_data, ggplot2::aes(x = BP_CUM, y = Y)) +
#     ggplot2::geom_point(ggplot2::aes(color = CHR), alpha = 0.8, size = 1.5) +
#     ggplot2::scale_color_manual(values = rep(c("grey30", "navy"), nlevels(plot_data$CHR))) +
#     ggplot2::scale_x_continuous(label = names(axis_breaks), breaks = axis_breaks) +
#     ggplot2::theme_minimal() +
#     ggplot2::theme(legend.position = "none",
#                   panel.grid.major.x = ggplot2::element_blank(),
#                   panel.grid.minor.x = ggplot2::element_blank()) +
#     ggplot2::labs(x = "Chromosome", y = expression(-log[10](italic(P))))
# }
#
# # Run GWAS
# GWAS_rrBLUP_Full <- rrBLUP::GWAS(pheno = pheno_file,
#                                  geno = marker_mat,
#                                  n.PC = 3,
#                                  plot = FALSE)
#
# # Plot
# plot_gwas(GWAS_rrBLUP_Full, chr = "chr", pos = "pos", y = "GY_predicted.value")
#
# # Run GWAS
# GWAS_rrBLUP_Sub <- rrBLUP::GWAS(pheno = pheno_file,
#                                 geno = marker_mat[marker_mat$chr %in% c("4A"),],
#                                 n.PC = 3,
#                                 plot = FALSE)
# #### Save ####
# save.image(file = "Final_Image.RData")
```